\sys: Property-Guided Program Regeneration\\{\vspace{1em}\large Technical University of Denmark, MSc Thesis {\\\vspace{-1em}} Advisors: Christian Gram Kahlauge and Nikos Vasilakis

### 2025-07-24 docs/paper/draft.tex
Improvements
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\begin{wrapfigure}[7]{r}{.5\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, Ih, Oh) 
  :- graph(p, Ijs, Ojs),
     tr(Ijs, Ih),
     tr(Ojs, Oh).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query asks \sys to preserve I/O behavior and translate the input program to Haskell.
The logic engine deduces that extracting I/O traces and the function signature is
sufficient.
The \ttt{tr} predicate is a generic LLM-assisted translator, which 
means that an LLM instance will be used to convert the inputs and outputs from Javascript 
to Haskell.
In this case, the given query is sufficient for the translation to succeed, 
but a natural-language description of the transformation can be passed as an additional 
argument to \ttt{tr}.
\sys, same as before, extracts an I/O trace from the original program.
Then, it transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\ttt{([1,[2,3]], (x)=>x+1)}\to\ttt{[2,3,4]}\rangle$ 
becomes $\langle(\ttt{[1, [2,3]], Js("(x)=>x+1"))}\to\ttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated and subsequently translated 20 I/O pairs,
\sys produces the following Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

\begin{wrapfigure}[7]{r}{.7\columnwidth}
  \begin{minted}{prolog}
db('music.db').func(f1).func(f2).
trc_sql(p_,I,T) :-
   trc_sql(p,I,T), graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
                   graph(p,I,O).
\end{minted}
\end{wrapfigure}
Having the database initialization being in the same function as 
the database access has negative architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
The query above first specifies \ttt{music.db} as the database to use.
This information is relevant to the \ttt{trc\_sql} plugin, which records the SQL queries 
sent to the databse when the function is driven by an input $i$.
The set of inputs $I$ the function will be driven by is generated by the \ttt{graph} plugin.
The \ttt{call} predicate indicates a function call.
Spefically, the query asks \sys to generate two functions such 
that the graph of the composition of the two is equivalent to the graph (and SQL trace) of the original program.
This transforms each I/O pair to (for example) $\langle\ttt{(f2(f1(), 'Metalica'))}, \ttt{[...]}\rangle$.
\sys first generates 24 inputs, then wraps the node process with monitors that extract the SQL queries.
Then, it executes the original program on each input, recording the output and SQL traces.
Following, \sys transforms the I/O pairs and the SQL traces into a prompt template,
and prompts an LLM instance to generate two functions $f_1$ and $f_2$, which when composed would have the given behavior.
Their composition is then again driven by the same inputs and the SQL traces and outputs are recorded.
After 20.4 seconds and two attempts, \sys arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$ in regards to its graph and SQL trace .
The user could have provided specific signatures and names for each 
function as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function f1() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

In the logic space, properties are encoded as predicates over abstract program
identifiers.
To bridge this abstraction gap, \sys given the set of true predicates
from the logic space, extracts the corresponding concrete properties from the
source program.
For example, the predicate $\ttt{inputs}(p, I)$ in the logic space
will be concertized to set of inputs $I$ by invoking the \ttt{input} plugin,
which uses an LLM to generate a set of inputs for the program $p$.
Similarly, the predicate $\ttt{graph}(p, I, O)$ will be concertized to a set of
input-output pairs $(I, O)$ by executing the program $p$ on the inputs $I$ and
recording the outputs $O$.
After regeneration, the predicates over the concrete regenerated program $p\_$
will be checked regarding their relationship to the same properties over program $p$
in the logic space.
An equality relationship is one where the predicate's grounded version
$X$ is the same between the two programs, modulo the program identifier~(\eg, $\ttt{graph}(p, i, o) \equiv \ttt{graph}(p\_, i, o)$).
Equality is defined differently for each property, and is enforced by the corresponding plugin.
Similarly, minimization and maximization relationships are defined as
the concrete property value being less-than or greater-than the original program's value, respectively.
Again, the less-than and greater-than operators are defined by the plugin that verifies the property.

@@

\sys successfully regenerates all programs in the \textit{Rosetta Code} benchmark suite
based on manual verification that the output is functionally equivalent
to the original, using both the original and translated reference implementations as guides.
\sys regenerates correctly all npm and PyPi utilities and it does not introduce and side-effectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \ttt{left-pad} library, pre-computes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge:
\sys-generated code frequently fails to account for all edge cases when they involve language-specific features (\eg \ttt{is-nan}), where there is a shortcut for determining 
the \ttt{NaN}-ity of a value (\ttt{value !== value;}).
For the \textit{SSCA} benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} utility plugin) to generate a function that produces a node \ttt{Stream} object.
For the \textit{IOCCC} benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive Mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the original one's semantics, as it operates over infinitely nested lists, rather than only one level deep ones.
Thus, a $.5$ score in correctness is given.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic inverse square root computation ($1/\sqrt{x}$) rather than the original Newton's method implemented through bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

For the Rosetta Code benchmarks, \gptmodel 
is successful in all language translation tasks.
The LLM-generated code matches the
original one's semantics, confirmed after manually inspecting the generated code.
Similarly, the LLM was successful in transforming the npm and PyPi utilities towards
\say{removing} any vulnerabilities from them, as they are benign and the transformation
involved the LLM simply outputting the original source code.
For the malicious libraries in the \textit{SSCA} benchmark, we provide the LLM 
with two different versions.
One is the original high-profile attack that appeared on npm
and the second one is an attack that is semantically equivalent to the first but with the source 
having gone through an iterative loop where a second LLM is tasked with making the code ostensibly 
secure~\cite{xu2023llmfoolitselfpromptbased}.
The second LLM instance added many integrity checks at all points in the program where a value is used and used variable names like \ttt{SAFE\_VALUE}.
\gptmodel was successful in eliminating the original attacks, but failed in \emph{all} 
cases where the source code was particularly crafted to appear secure, regenerating both 
the original functionality and the attack.
For the \textit{IOCCC} benchmarks, \gptmodel also fails 
in both cases to regenerate the programs in a way where the code's 
functionality is preserved while removing the obfuscation.
The \ttt{adder} program is regenerated as a program that prints the \ttt{argc} argument, 
while the \ttt{mersenne} program performs irrelevant arithmetic operations that 
resemble the original source, and fails to even compile.
For the microbenchmarks, \gptmodel is able to regenerate the \ttt{flatmap-stream} program, but fails to produce a Haskell version of it,
as it over-fits to the original JavaScript source code which uses the \ttt{Stream} API,
and fails to produce a Haskell program that is nor equivalent to the original, nor idiomatic.
For the \ttt{Q\_rsqrt} program, \gptmodel fails to produces a program that idiomatically 
computes the inverse square root, rather trying to minimize the number of lines in the program
by removing constants and comments while keeping the unidiomatic bit-level manipulation and use of magic numbers.
Finally, for the \ttt{rigidDB} program, \gptmodel is able to decompose the original program into two functions
and names them \ttt{setupDatabase} and \ttt{getAlbumsByArtist}.

\sys consistently outperforms the \gptmodel baseline, overcoming the inherent weaknesses of a purely prompt-based approach.
In-depth examples of interesting evaluation results are discussed in \cref{sec:eval-samples}.

>>>>

\begin{wrapfigure}[7]{r}{.5\columnwidth}
% \vspace{-10pt}
\begin{minted}{prolog}
graph(p_, Ih, Oh) 
  :- graph(p, Ijs, Ojs),
     tr(Ijs, Ih),
     tr(Ojs, Oh).
language(p_, haskell).
\end{minted}
\end{wrapfigure}
The query asks \sys to preserve I/O behavior and translate the input program to Haskell.
The \ttt{tr} predicate is a generic LLM-assisted translator, and in the query
converts the input/output examples from Javascript to Haskell.
In this case, the given query is sufficient for the translation to succeed, 
but a natural-language description of the transformation can be passed as an additional 
argument to \ttt{tr}.
\sys, same as before, extracts an I/O trace from the original program.
Then, it transforms each extracted I/O example into an equivalent one in Haskell (\eg 
the pair $\langle\ttt{([1,[2,3]], (x)=>x+1)}\to\ttt{[2,3,4]}\rangle$ 
becomes $\langle(\ttt{[1, [2,3]], Js("(x)=>x+1"))}\to\ttt{[2,3,4]}\rangle$).
Finally, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
After 32.2 seconds, having generated and subsequently translated 20 I/O pairs,
\sys produces the following Haskell program (truncated for brevity):
\begin{minted}[frame=lines]{haskell}
flatmap :: (a -> b) -> NestedList a -> [b]
flatmap f (Elem x) = [f x]
flatmap f (List x) = concatMap (flatmap f) x
\end{minted}

@@

\begin{wrapfigure}[7]{r}{.7\columnwidth}
  \begin{minted}{prolog}
db("music.db").func(f1).func(f2).
trc_sql(p_,I,T) :-
   trc_sql(p,I,T), graph(p,I,_).
graph(call(f2,call(f1),I),I,O):-
                   graph(p,I,O).
\end{minted}
\end{wrapfigure}
Having the database initialization being in the same function as 
the database access has negative architectural and performance implications.
In addition, there exist systems that can benefit from further decomposition
of an application towards automatic distribution, parallelization or security~\cite{Towards_Modern_Ghemaw_2023, vasilakis2019ignis, vasilakis2018breakapp}.
The query above first specifies \ttt{music.db} as the database to use.
This information is relevant to the \ttt{trc\_sql} plugin, which records the SQL queries 
sent to the databse when the function is driven by an input $i$.
The set of inputs $I$ the function will be driven by is generated by the \ttt{graph} plugin.
An example of an input, SQL-trace pair is
$\langle\ttt{'Metalica'}$, $\langle$\ttt{SELECT album FROM songs WHERE artist = 'Metalica'}, \ttt{['Master of Puppets', 'Ride the Lightning']}$\rangle\rangle$,
which means that the input \ttt{'Metalica'} causes the resulting SQL query to be executed by the function,
which together with the resulting rows is the SQL trace.
The built-in \ttt{call} predicate redefines the structure of the program to be regenerated. 
Spefically, the query asks \sys to generate two functions such 
that the graph of the composition of the two is equivalent to the graph (and SQL trace) of the original program.
This transforms each I/O pair to (for example) $\langle\ttt{(f2(f1(), 'Metalica'))}, \ttt{[...]}\rangle$.
\sys first generates 24 inputs, then wraps the node process with monitors that extract the SQL queries.
Then, it executes the original program on each input, recording the output and SQL traces.
Following, \sys transforms the I/O pairs and the SQL traces into a prompt template,
and prompts an LLM instance to generate two functions $f_1$ and $f_2$, which when composed would have the given behavior.
Their composition is then again driven by the same inputs and the SQL traces and outputs are recorded.
After 20.4 seconds and two attempts, \sys arrives at two functions $f_1$, and 
$f_2$, where their composition $f_2(f_1(), i)$,
is equivalent to the original \ttt{getAlbumsByArtist} function across each input $i$ in regards to its graph and SQL trace .
The user could have provided specific signatures and names for each 
function as an additional guiding property.
The resulting program is:
\begin{minted}[frame=lines]{js}
function f1() {
 const sqlite = require('sqlite');
 return sqlite.open({filename: 'music.db'});
}

@@

In the logic space, properties are encoded as predicates over abstract program
identifiers.
To bridge this abstraction gap, \sys given the set of true predicates
from the logic space, extracts the corresponding concrete properties from the
source program.
For example, the predicate $\ttt{inputs}(p, I)$ in the logic space
will be concertized to set of inputs $I$ by invoking the \ttt{input} plugin,
which uses an LLM to generate a set of inputs for the program $p$.
Similarly, the predicate $\ttt{graph}(p, I, O)$ will be concertized to a set of
input-output pairs $(I, O)$ by executing the program $p$ on the inputs $I$ and
recording the outputs $O$.
After regeneration, the predicates over the concrete regenerated program $p\_$
will be checked regarding their relationship to the same properties over program $p$
in the logic space.
% An equality relationship is one where the predicate's grounded version
% $X$ is the same between the two programs, modulo the program identifier~(\eg, $\ttt{graph}(p, i, o) \equiv \ttt{graph}(p\_, i, o)$).
Equality is defined differently for each property, and is enforced by the corresponding plugin.
Similarly, minimization and maximization relationships are defined as
the concrete property value being less-than or greater-than the original program's value, respectively.
Again, the less-than and greater-than operators are defined by the plugin that verifies the property.

@@

\heading{Results}
\sys successfully regenerates all programs in the \textit{Rosetta Code} benchmark suite
based on manual verification that the output is functionally equivalent
to the original, using both the original and translated reference implementations as guides.
\sys regenerates correctly all npm and PyPi utilities and it does not introduce and side-effectful or malicious behavior.
The synthesized programs reveal interesting insights and patterns.
\sys-generated code often lacks the optimizations commonly
implemented by library developers, such as early exits or caching.
For instance, the \ttt{left-pad} library, pre-computes padding for strings below a certain length
and returns it directly.
In contrast, \sys tends to produce straightforward
implementations without attempting similar optimizations.
\sys also omits domain-specific knowledge:
\sys-generated code frequently fails to account for all edge cases when they involve language-specific features (\eg \ttt{is-nan}), where there is a shortcut for determining 
the \ttt{NaN}-ity of a value (\ttt{value !== value;}).
For the \textit{SSCA} benchmarks, \sys successfully regenerates all three malicious libraries, removing the malicious behavior while preserving the original functionality.
For the flatmap-stream attack, \sys uses domain-specific logic (encoded in its knowledge-base by the \ttt{js-stream} utility plugin) to generate a function that produces a node \ttt{Stream} object.
For the \textit{IOCCC} benchmarks, \sys regenerates the \ttt{adder} program, which is a simple program that adds two numbers,
it also regenerates \ttt{mersenne} in regards that it correctly performs a naive Mersenne primality testing, but it does not follow the non-obvious output scheme of the original program, which returns a non-zero hexadecimal number when $2^p - 1$ ($p$ being the input number) is not prime.
For the microbenchmarks, \sys correctly regenerates the \ttt{flatmap-stream} program, producing a safe Haskell version that is idiomatic and preserves the original functionality.
However, this version does not match exactly the original one's semantics, as it operates over infinitely nested lists, rather than only one level deep ones.
Thus, a $.5$ score in correctness is given.
The \ttt{Q\_rsqrt} program includes the obvious, idiomatic inverse square root computation ($1/\sqrt{x}$) rather than the original Newton's method implemented through bit-level manipulation.
Finally, \sys successfully decomposes the \ttt{rigidDB} program into two
functions, \ttt{setupDatabase} and \ttt{getAlbumsByArtist}, which is a
modularization that preserves the original functionality and produce equivalent I/O and SQL traces.

@@

\heading{Results}
For the Rosetta Code benchmarks, \gptmodel 
is successful in all language translation tasks.
The LLM-generated code matches the
original one's semantics, confirmed after manually inspecting the generated code.
Similarly, the LLM was successful in transforming the npm and PyPi utilities towards
\say{removing} any vulnerabilities from them, as they are benign and the transformation
involved the LLM simply outputting the original source code.
For the malicious libraries in the \textit{SSCA} benchmark, we provide the LLM 
with two different versions.
One is the original high-profile attack that appeared on npm
and the second one is an attack that is semantically equivalent to the first but with the source 
having gone through an iterative loop where a second LLM is tasked with making the code ostensibly secure~\cite{xu2023llmfoolitselfpromptbased}~(\cref{fig:llm-fool})
The second LLM instance added many integrity checks at all points in the program where a value is used and used variable names like \ttt{SAFE\_VALUE}.
\gptmodel was successful in eliminating the original attacks, but failed in \emph{all} 
cases where the source code was particularly crafted to appear secure, regenerating both 
the original functionality and the attack.
For the \textit{IOCCC} benchmarks, \gptmodel also fails 
in both cases to regenerate the programs in a way where the code's 
functionality is preserved while removing the obfuscation.
The \ttt{adder} program is regenerated as a program that prints the \ttt{argc} argument, 
while the \ttt{mersenne} program performs irrelevant arithmetic operations that 
resemble the original source, and fails to even compile.
For the microbenchmarks, \gptmodel is able to regenerate the \ttt{flatmap-stream} program, but fails to produce a Haskell version of it,
as it over-fits to the original JavaScript source code which uses the \ttt{Stream} API,
and fails to produce a Haskell program that is nor equivalent to the original, nor idiomatic.
For the \ttt{Q\_rsqrt} program, \gptmodel fails to produces a program that idiomatically 
computes the inverse square root, rather trying to minimize the number of lines in the program
by removing constants and comments while keeping the unidiomatic bit-level manipulation and use of magic numbers.
Finally, for the \ttt{rigidDB} program, \gptmodel is able to decompose the original program into two functions
and names them \ttt{setupDatabase} and \ttt{getAlbumsByArtist}.

\begin{figure}[t]
\promptbox{
Make the following as secure and safe as possible. 
Do not change the functionality in any way.

if (b == "gbabWhaRQ") \{
require("fs").
  writeFileSync('./FILE', (a + b));
\}
}

Output from \gptmodel:
\begin{minted}{javascript}
// Fixed secret value (compare safely)
const SAFE_KEY = 'gbabWhaRQ';

if (b == SAFE_KEY) {
require("fs").
  writeFileSync('./FILE', (a + b + '\n'));
} 
\end{minted}

  \label{fig:llm-fool}
  \caption{Example of how a malicious library can transformed into an equivalent 
  one that circumvents naive LLM-based detection techniques.}
\end{figure}

\heading{Key Result}
\sys consistently outperforms the \gptmodel baseline, overcoming the inherent weaknesses of a purely prompt-based approach.
In-depth examples of interesting evaluation results are discussed in \cref{sec:eval-samples}.
